\project Use the \lex\ and \yacc\ programs you have written for
\LaTeX\ to write a full \LaTeX-to-HTML translator.
%
%
%
\project A number of projects involve parsers or code generators for (parts of)
\TeX\ or \LaTeX.
\begin{description}
\item[formulas]
Reinhold Heckmann and Reinhard Wilhelm. 1997.  A functional
description of TeX's formula layout.  Journal of Functional
Programming 7(5):451-485. Available online at
\url{http://rw4.cs.uni-sb.de/users/heckmann/doc.html}.
For software, see \url{ftp://ftp.cs.uni-sb.de/formulae/}.
\item[also formulas]
Preview-LaTeX (\url{http://preview-latex.sourceforge.net/}) displays
formulas right in the emacs edit buffer.
\item[math on web pages]see \url{http://www.forkosh.com/mimetex.html}.
\item[\LaTeX\ to HTML] HeVeA, TtH, TeX4ht and LaTeX2HTML.
\item[front end for \LaTeX] \url{http://www.lyx.org/}
Ages ago there was `griff'. Scientific Word still exists but is commercial.
\item[reimplementation of \TeX] \TeX\ in Python: \url{http://www.pytex.org/}
\end{description}
Investigate these parsers: what is their basic approach and theoretical
power, what are they good at, what are they bad at.
%
%
%
\project Develop the full theory of the compound NFA that does lexical
analysis.
\begin{itemize}
\item This automaton basically parses the whole file, rather than
  small chunks; every once in a while it will report that it has
  recognized an identifier, number, special symbol~\&c. This means
  that the definition of the output alphabet has to be
  expanded. Analyze the structure of this output language.
\item As explained, the returning $\epsilon$-transition only has to be
  taken if a maximal string is recognized. Formalyze this notion.
\item The automaton may need look-ahead of one or more
  tokens. Formalize this notion and give equivalence proofs.
\end{itemize}
%
%
%
\project Do an experimental study of hashing. Implement open and
closed hash table algorithms, and do a search on the web for hash
functions. Compare to storing the names in a tree.
Find some documents to hash: the source of some programs,
books. Construct some input data sets to trip up the various
algorithms.
Measure statistics and do timings.
